Methods paper for erosion

### 2017-10-02 docs/methods/erosion_methods.tex
worked on numerics section of methods paper

\cite{moo-ris-chi-zha-she2013}

@@

\todo[inline]{Describe the erosion model and its justification}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
There are three main numerical methods that need to developed: the
boundary integral equation solver~\eqref{eqn:completed_BIE}, computing
the quantities of interest in Section~\ref{sec:qois}, and updating the
shape of the eroding bodies with equation~\eqref{eqn:bodyEvolution}.  In
order to compute stable simulations with long time horizons, we use
numerical methods that achieve spectral accuracy in space, second-order
in time, and maintain an equispaced discretization of the bodies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
Each body will be represented with $N_\iin$ points.  Because
the boundary of each body is closed, it can be represented with spectral
accuracy as a Fourier series
\begin{align*}
  \xx^k(\theta) = \sum_{j = -N/2}^{N/2-1} \hat{x}_j^k e^{ij\theta}.
\end{align*}
There are several instances where derivatives of the shape must be
taken, and this is done with spectral accuracy by differentiating the
Fourier series term-by-term
\begin{align*}
  \frac{d\xx^k(\theta)}{d\theta} = 
      \sum_{j = -N/2}^{N/2-1} ij\hat{x}_j^k e^{ij\theta}.
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature} 
There are several integrals that need to be evaluated.  This includes
the double-layer potential for target points both
off~\eqref{eqn:velocityDLP} and on~\eqref{eqn:DLP} the $\bd\Omega$, the
shear stress~\eqref{eqn:shearStressDLP}, the
vorticity~\eqref{eqn:vorticityDLP}, the pressure both
off~\eqref{eqn:pressureDLP} and on~\eqref{eqn:pressureDLPdiag} the
boundary, the drag~\eqref{eqn:drag}, and the resistivity.  Most of the
integrands involved in these calculations are both smooth and periodic,
and the suitably adjusted trapezoid rule will result in spectral
accuracy~\cite{tre-wei2014}.

To numerically solve the completed double-layer
potential~\eqref{eqn:completed_BIE}, we start by discretizing each
interior body $\gamma_k$ with $N_\iin$ points, and the outer geometry
$\Gamma$ with $N_\out$ points.  We let $N = MN_\iin + N_\out$ be the
total number of discretization points, and we apply the trapezoid rule
which results in the linear system
\begin{align*}
  \ff(\xx_i^k) = \sum_{\ell = 1}^{M}\sum_{j=1}^{N} 
    K(\xx_i^k,\xx_j^\ell) \eeta(\xx_j^\ell) 
      \Delta s_{j}^{\ell}, \quad i=1,\ldots,N, \: k=1\ldots,M
\end{align*}
where $\Delta s_j^\ell$ is the arclength term of $\gamma_\ell$ at
$\xx_j^\ell \in \gamma_\ell$, or the corresponding term if $\xx_k \in
\Gamma$, and
\begin{align*}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2},
\end{align*}
where $\rr = \xx - \yy$, and $\rho = \|\rr\|$, is the kernel of the
double-layer potential.  The Stokeslet~\eqref{eqn:stokeslet} and
rotlet~\eqref{eqn:rotlet} strengths are also discretized with the
trapezoid rule
\begin{align*}
  \llambda_k &= \sum_{k=1}^{N_\iin} \eeta(\xx_k) 
    \Delta s_{k} \quad \text{and} \quad
  \mu_k = \sum_{k=1}^{N_\iin} \xx_k^\perp \cdot
    \eeta(\xx_k) \Delta s_{k}.
\end{align*}
\todo[inline]{need to pick a good notation for the indices}

By applying a quadrature method, this results in a dense linear system
which we solve with fast multipole accelerated GMRES.  Because of the
second-kind nature of the integral equation, the number of GMRES
iterations is mesh independent (but not geometry independent), and the
overall complexity of the Stokes solver is proportional to the cost of a
single matrix-vector multiplication.  The shear stress, pressure, and
drag must also be computed using quadrature, and this can be interpreted
as a post-processing step once the density function, Stokeslets, and
rotlets are computed.  All the quadrature methods are described in
Section~\ref{sec:quadrature}

@@

\label{sec:quadrature}
There are several integrals that we need to compute including the
double-layer potential velocity field~\eqref{eqn:velocityDLP}, the shear
stress~\eqref{eqn:shearStressDLP}, vorticity~\eqref{eqn:vorticityDLP},
and the pressure~\eqref{eqn:pressureDLP}.  By using carefully choosing
the quadrature rules, high-order solutions can be computed.  For all the
integrals, the trapezoid rule is used since it achieves spectral
accuracy for smooth periodic functions.  However, care must be taken
when the source point and target points coincide.

For the velocity field, the kernel of the integral operator
\begin{align*}
  D(\xx,\yy) = \frac{\rr \cdot \nn}{\rho^2} 
               \frac{\rr \otimes \rr}{\rho^2}
\end{align*}
has a removable singularity when $\xx = \yy$.  In particular
\begin{align*}
  \lim_{\substack{\yy \rightarrow \xx \\ \yy \in \bd\Omega}} D(\xx,\yy) = 
  -\frac{\kappa(\xx)}{2\pi}\ss \otimes \ss,
\end{align*}
where $\kappa(\xx)$ is the curvature of $\bd\Omega$ at $\xx$.  By using
this limiting term in when applying the trapezoid rule, spectral
accuracy is achieved because of the smoothness and periodicity of the
integrand~\cite{tre-wei2014}.

For the shear stress and the pressure, we will use odd-even integration
which achieves spectral accuracy if the integrand is no stronger than
$\mathcal{O}(\rho^{-1})$.  Since both these integrands are too strong,
the singularity must be weakened.  This can be done by adding a constant
density function which has no effect on the shear stress and
pressure~\cite{poz1992}.  In particular, the shear stress due to the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx_0) &= \int_{\bd\Omega}K_\ssigma(\xx_0,\yy)
      \ssigma(\yy) \, ds_{\yy} \\ 
      &= \int_{\bd\Omega}K_\ssigma(\xx_0,\yy)
      (\ssigma(\yy) - \ssigma(\xx_0)) \, ds_{\yy}.
\end{align*}
For the pressure, the double-layer potential is equivalent to
\begin{align*}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx_0)) \,ds_\yy.
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fast Summation} 
The majority of the computational cost comes from evaluating the the
double-layer potential $\mathcal{D}$ in equation~\eqref{eqn:DLP}.  This
layer potential can be efficiently evaluated using a fast summation
method such as the Fast Multipole Method
(FMM)~\cite{gre-rok1987}\footnote{The FMM for the Stokes double-layer
poetntial was supplied by Manas Rachh.}.  By using the
FMM,~\eqref{eqn:completed_BIE} can be solved with $\mathcal{O}(N)$
operations, where $N$ is the total number of discretization points.

In our erosion model, the shear stress must be computed which requires
evaulating the layer potential~\eqref{eqn:shearStressDLP}.  In our
implementation, this calculation is done directly which requires
$\mathcal{O}(N^2)$ operations.  This complexity could be reduced with
the kernel-independent FMM~\cite{yin-bir-zor2004}.  However, since this
calculation is only computed once per time step, unlike the velocity
double-layer potential that must be evaluated at each GMRES iteration.

>>>>

@@

\todo[inline]{Describe the erosion model and its justification}
\cite{moo-ris-chi-zha-she2013}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical Methods\label{s:method}} 
There are three main numerical methods that need to be developed:~the
boundary integral equation solver~\eqref{eqn:completed_BIE}, computing
the quantities of interest in Section~\ref{sec:qois}, and updating the
shape of the eroding bodies with equation~\eqref{eqn:bodyEvolution}.  In
order to compute stable simulations with long time horizons, we use
numerical methods that achieve spectral accuracy in space, second-order
in time, and maintain an equispaced discretization of the bodies.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spatial Discretization}
Each body will be represented with $N_\iin$ points.  The boundary is
represented with the Fourier series
\begin{align*}
  \xx^k(\theta) = \sum_{j = -N/2}^{N/2-1} \hat{x}_j^k e^{ij\theta}.
\end{align*}
Derivatives of the shape must be taken which is easily done by
differentiating the Fourier series term-by-term
\begin{align*}
  \frac{d\xx^k(\theta)}{d\theta} = 
      \sum_{j = -N/2}^{N/2-1} ij\hat{x}_j^k e^{ij\theta}.
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature} 
Integrals that need to be approximated are the double-layer potential
for target points both off the boundary~\eqref{eqn:velocityDLP} and on
the boundary~\eqref{eqn:DLP}, the shear
stress~\eqref{eqn:shearStressDLP}, the
vorticity~\eqref{eqn:vorticityDLP}, the pressure both off the
boundary~\eqref{eqn:pressureDLP} and on the
boundary~\eqref{eqn:pressureDLPdiag}, the drag~\eqref{eqn:drag}, and the
resistivity.  The integrands of each of these integrals, except for the
resistivity, are smooth and periodic.  Under this condition the
trapezoid rule guarantees spectral accuracy~\cite{tre-wei2014}.
However, before the trapezoid rule can be applied, singularity
subtraction and odd-even integration~\cite{sid-isr1988} method must be
used.

\subsubsection{Velocity}
To numerically evaluate the completed double-layer
potential~\eqref{eqn:completed_BIE}, we start by discretizing each
interior body $\gamma_k$ with $N_\iin$ points, the outer geometry
$\Gamma$ with $N_\out$ points, and define $N = MN_\iin + N_\out$ to be
the total number of discretization points.  The double-layer potential
in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i^k) = \sum_{\ell = 1}^{M}\sum_{j=1}^{N} 
    K(\xx_i^k,\xx_j^\ell) \eeta(\xx_j^\ell) 
      \Delta s_{j}^{\ell}, \quad i=1,\ldots,N, \: k=1\ldots,M,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j^\ell$ is the arclength term of $\gamma_\ell$ at
$\xx_j^\ell \in \gamma_\ell$, or the corresponding term if $\xx_k \in
\Gamma$, and
\begin{align*}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2},
\end{align*}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the
double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is
replaced with the limiting value of the double-layer potential
\begin{align*}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)),
\end{align*}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent
vector at $\xx \in \bd\Omega$.  The Stokeslet~\eqref{eqn:stokeslet} and
rotlet~\eqref{eqn:rotlet} strengths are also discretized with the
trapezoid rule, and the resulting discretization
of~\eqref{eqn:completed_BIE} is
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i^k) &= -\frac{1}{2}\ssigma(\xx_i^k) + 
      \sum_{\ell = 1}^{M}\sum_{j=1}^{N} 
      K(\xx_i^k,\xx_j^\ell) \eeta(\xx_j^\ell) \Delta s_{j}^{\ell} +
      \sum_{\ell=1}^M S[\llambda_k,\cc_k](\xx_i^k) +
      \sum_{\ell=1}^M R[\xi_k,\cc_k](\xx_i^k), \\
    \llambda_k &= \sum_{i=1}^{N_\iin} \eeta(\xx_i^k) 
      \Delta s_i^k \\ 
    \mu_k &= \sum_{i=1}^{N_\iin} {(\xx_i^k)}^\perp \cdot
      \eeta(\xx_i^k) \Delta s_i^k,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $k=1,\ldots,M$.

The dense linear system~\eqref{eqn:completed_BIE_discretized} is solved
iteratively with GMRES.  The most expensive part is computing the
double sum for all target and source points.  A direct summation would
require $\bigO(N^2)$ operations, but this is reduced to $\bigO(N)$ by
using the Fast Mutlipole method for the Stokes double-layer
potential~\cite{gre-rok1987,rac-???} \todo[inline]{sent email to Manas
asking about this}.  Since we are discretizing a second-kind integral
equation, the number of GMRES iterations is bounded independent of
$N$~\cite{cam-ips-kel-mey-xue1996}, and the overall complexity for
solving~\eqref{eqn:completed_BIE_discretized} is $\bigO(N)$.

\subsubsection{Shear Stress}
In our erosion model, the shear stress must be computed which requires
evaulating the layer potential~\eqref{eqn:shearStressDLP}.  In our
implementation, this calculation is done directly which requires
$\bigO(N^2)$ operations.  This complexity could be reduced with
the kernel-independent FMM~\cite{yin-bir-zor2004}.  However, since this
calculation is only computed once per time step, unlike the velocity
double-layer potential that must be evaluated at each GMRES iteration.

\subsubsection{Vorticity}

\subsubsection{Pressure}

\subsubsection{Drag}

\subsubsection{Resistivity}

The shear stress, pressure, and drag must also be computed using
quadrature, and this can be interpreted as a post-processing step once
the density function, Stokeslets, and rotlets are computed.  All the
quadrature methods are described in Section~\ref{sec:quadrature}

@@

For the shear stress and the pressure, we will use odd-even integration
which achieves spectral accuracy if the integrand is no stronger than
$\bigO(\rho^{-1})$.  Since both these integrands are too strong, the
singularity must be weakened.  This can be done by adding a constant
density function which has no effect on the shear stress and
pressure~\cite{poz1992}.  In particular, the shear stress due to the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx_0) &= \int_{\bd\Omega}K_\ssigma(\xx_0,\yy)
      \ssigma(\yy) \, ds_{\yy} \\ 
      &= \int_{\bd\Omega}K_\ssigma(\xx_0,\yy)
      (\ssigma(\yy) - \ssigma(\xx_0)) \, ds_{\yy}.
\end{align*}
For the pressure, the double-layer potential is equivalent to
\begin{align*}
  p^\DD(\xx) = -\frac{1}{\pi}\int_{\bd\Omega} \frac{1}{\rho^2}
    \left(I - 2 \frac{\rr \otimes \rr}{\rho^2}\right) 
    \nn \cdot (\eeta(\yy) - \eeta(\xx_0)) \,ds_\yy.
\end{align*}
